---
title: "11 Data Import"
author: "Russ Conte"
date: "9/1/2018"
output: html_document
---

```{r}
library(tidyverse)
heights <- read_csv(file =  'https://raw.githubusercontent.com/hadley/r4ds/master/data/heights.csv') #yay!
heights

```

Note it is possible to read an inline CSV file:

```{r}
read_csv("a,b,c # note no commat at the end of each ROW of data
         1,2,3
         4,5,6")
```

looking at the newline command and NA:

```{r}
read_csv("a,b,c\n1,2,3\n4,na,6", na=".") # note first line is column names
```

Exercises

1. What funciton would you use to read a file where fields are separated with "|"?
read_delim and then set the delimiter to "|"

2. Apart from file, skip and comment, what other arguments to read_csv and read_tsv have in common?

read_csv(file, col_names = TRUE, col_types = NULL,
  locale = default_locale(), na = c("", "NA"), quoted_na = TRUE,
  quote = "\"", comment = "", trim_ws = TRUE, skip = 0, n_max = Inf,
  guess_max = min(1000, n_max), progress = show_progress())
  
read_tsv(file, col_names = TRUE, col_types = NULL,
  locale = default_locale(), na = c("", "NA"), quoted_na = TRUE,
  quote = "\"", comment = "", trim_ws = TRUE, skip = 0, n_max = Inf,
  guess_max = min(1000, n_max), progress = show_progress())
  
Common elements:
col_names=TRUE
col_types = NULL
locale = default_locale()
na=c("", "NA")
quoted_na = TRUE
quote = "\"
comment = ""
trim_ws=TRUE
skip = 0
n_max = inf
guess_max = min(1000, nmax)
progress = show_progress()

3. What are the most important arguments to read_fwf?
the file itself, columns positions and types.

4.

5. What is wrong with each of the following inline CSV files?

```{r}
read_csv("a,b,\n1,2,3\n4,5,6") # note that column 3 is not named, so R gives it the name X3
```

```{r}
read_csv("a,b,c\n1,2\n1,2,3,4") #unequal number of columns in the data as parsed results in NA
```

Another example with an NA.

## parsing a vector

Note - parsing a vector converts into another type of vector, such as a logical, integer, string, etc.

Example: Convert string to logical
```{r}
str(parse_logical(c("TRUE", "FALSE", "NA")))
```

Convert string to interger
```{r}
str(parse_integer(c("1", "2", "3")))
```

Convert character to actual date format:
```{r}
str(parse_date(c("2018-09-02", "0100-12-25")))
```

parse interger:
```{r}
parse_integer(c("1","333",".","456"),na=".") # note the NA goes OUTSIDE of the data
```

If parsing failes, R generates an error:

```{r}
x <- parse_integer(c("1","2","abc","."), na=".")
```

It's possible to use problems() to address parsing issues:
```{r}
problems(x)
```

## 11.3.1 Parsing numbers

There exist three issues with numbers:
1. People in different parts of the world write the same numbers differently:

12,38 or 12.38, for example

```{r}
parse_double(c("12.38"))
parse_double(c("12,30")) # returns an error because R (my version) is coded for the USA.
```

2. Numbers are often surrounded by symbols:
$40.00, 45%, -$5.00

parse_number handles this quite easily:
```{r}
parse_number(c("$1000"))
parse_number(c("$345.78"))
parse_number(c("45%"))
parse_number(c("We had 78 people in attendance")) # wow! This is amazing! Wowsie!
```



3. Numbers often contain grouping characters that make them easier for humans to read, but more difficult for computers:
123,456,789

```{r}
parse_number(c("123,456,789"))
```

## 11.3.2 Strings

```{r}
x1 <- "El Ni\xf1o was particularly bad this year"
x2 <- "\x82\xb1\x82\xf1\x82\xc9\x82\xbf\x82\xcd"

x1
x2
```

To fix these issues, put the encoding into the statement:

```{r}
parse_character(x1,locale = locale(encoding = "Latin1"))
parse_character(x2, locale = locale(encoding = "Shift-JIS"))
```

It's possible to have R guess the encoding, here's what that looks like:

```{r}
guess_encoding(charToRaw(x2)) # it guessed OI8-R, so let's try that one:
parse_character(x2, locale=locale(encoding = "KOI8-R"))
```

##11.3.3 Factors

```{r}
fruit <- c("Apple", "Orange", "Banana")
parse_factor(c("Apple", "Orange", "Banana"),levels = fruit)
```

## 11.3.4 Dates, date-time and times

This will parse date-time according to ISO 8601

```{r}
parse_datetime(c("2018-09-02T1310"))
```

To parse time, we will use the hms package:

```{r}
library(hms)
parse_time("15:45")

```

## 11.3.5 Exercises