---
title: "11 Data Import"
author: "Russ Conte"
date: "9/1/2018"
output: html_document
---

## 11.2 Getting Started

From the text: "When you run read_csv() it prints out a column specification that gives the name and type of each column. That’s an important part of readr, which we’ll come back to in parsing a file."

```{r}
library(tidyverse)
heights <- read_csv(file =  'https://raw.githubusercontent.com/hadley/r4ds/master/data/heights.csv') #yay!
heights

```

Note it is possible to read an inline CSV file:

```{r}
# note no comm at at the end of each ROW of data
read_csv("a,b,c 
         1,2,3
         4,5,6")
```

Here is how to skip the top rows if they do not have data:

```{r}
read_csv("The first line of metadata
         The second line of metadata
         x,y,z
         1,2,3", skip=2)

read_csv("do not import this line,
         a,b,c
         4,5,6", skip=1)

```

If the first row does NOT have column names, here is how to tell R to NOT treat the first row as column names. R will create column names.

```{r}
read_csv("x,y,z\n1,2,3",col_names = FALSE)
```

It's possible to assign column names, too. Here's how that is done:

```{r}
read_csv("x,y,z\n1,2,3",col_names = c("First", "Second", "Third"))
```

It' virtually guarantted that data will have NAs, here is one way to deal with NAs.
This solution treats the period as an NA:

```{r}
read_csv("a,b,c\n1,2,3\n4,na,6", na=".") # note first line is column names
```

Exercises

1. What funciton would you use to read a file where fields are separated with "|"?
read_delim and then set the delimiter to "|", thus:

```{r}
library(readr)
read_delim("1|2|3\nx|y|z",delim="|")
```

2. Apart from file, skip and comment, what other arguments to read_csv and read_tsv have in common?

read_csv(file, col_names = TRUE, col_types = NULL,
  locale = default_locale(), na = c("", "NA"), quoted_na = TRUE,
  quote = "\"", comment = "", trim_ws = TRUE, skip = 0, n_max = Inf,
  guess_max = min(1000, n_max), progress = show_progress())
  
read_tsv(file, col_names = TRUE, col_types = NULL,
  locale = default_locale(), na = c("", "NA"), quoted_na = TRUE,
  quote = "\"", comment = "", trim_ws = TRUE, skip = 0, n_max = Inf,
  guess_max = min(1000, n_max), progress = show_progress())
  
Common elements:
col_names=TRUE
col_types = NULL
locale = default_locale()
na=c("", "NA")
quoted_na = TRUE
quote = "\"
comment = ""
trim_ws=TRUE
skip = 0
n_max = inf
guess_max = min(1000, nmax)
progress = show_progress()

3. What are the most important arguments to read_fwf?
the file itself, columns widths/positions and types.

```{r}
cat(read_lines(fwf_sample))
read_fwf(fwf_sample, fwf_empty(fwf_sample, col_names = c("first", "last", "state", "ssn")))

```


4.Sometimes strings in a CSV file contain commas. To prevent them from causing problems they need to be surrounded by a quoting character, like " or '. By convention, read_csv() assumes that the quoting character will be ", and if you want to change it you’ll need to use read_delim() instead. What arguments do you need to specify to read the following text into a data frame?
"x,y\n1,'a,b'"

Use the 'quote' function:

```{r}
read_csv("x,y\n1,'a,b'", quote = "'")
```


5. What is wrong with each of the following inline CSV files?

```{r}
read_csv("a,b,\n1,2,3\n4,5,6") # note that column 3 is not named, so R gives it the name X3
```

```{r}
read_csv("a,b,c\n1,2\n1,2,3,4") #unequal number of columns in the data as parsed results in NA
```

Another example with an NA.

## 11.3 parsing a vector

Note - parsing a vector converts into another type of vector, such as a logical, integer, string, etc.

Example: Convert string to logical
```{r}
str(parse_logical(c("TRUE", "FALSE", "NA")))
```

Convert string to interger
```{r}
str(parse_integer(c("1", "2", "3")))
```

Convert character to actual date format:
```{r}
str(parse_date(c("2018-09-02", "0100-12-25")))
```

parse interger:
```{r}
parse_integer(c("1","333",".","456"),na=".") # note the NA goes OUTSIDE of the data
```

If parsing failes, R generates an error:

```{r}
x <- parse_integer(c("1","2","abc","."), na=".")
```

It's possible to use problems() to address parsing issues:
```{r}
problems(x)
```

## 11.3.1 Parsing numbers

There exist three issues with numbers:
1. People in different parts of the world write the same numbers differently:

12,38 or 12.38, for example

```{r}
parse_double(c("12.38"))
parse_double(c("12,30")) # returns an error because R (my version) is coded for the USA.
```

2. Numbers are often surrounded by symbols:
$40.00, 45%, -$5.00

parse_number handles this quite easily:
```{r}
parse_number(c("$1000"))
parse_number(c("$345.78"))
parse_number(c("45%"))
parse_number(c("We had 78 people in attendance")) # wow! This is amazing! Wowsie!
```



3. Numbers often contain grouping characters that make them easier for humans to read, but more difficult for computers:
123,456,789

```{r}
parse_number(c("123,456,789"))
```

## 11.3.2 Strings

```{r}
x1 <- "El Ni\xf1o was particularly bad this year"
x2 <- "\x82\xb1\x82\xf1\x82\xc9\x82\xbf\x82\xcd"

x1
x2
```

To fix these issues, put the encoding into the statement:

```{r}
parse_character(x1,locale = locale(encoding = "Latin1"))
parse_character(x2, locale = locale(encoding = "Shift-JIS"))
```

It's possible to have R guess the encoding, here's what that looks like:

```{r}
guess_encoding(charToRaw(x2)) # it guessed OI8-R, so let's try that one:
parse_character(x2, locale=locale(encoding = "KOI8-R"))
```

##11.3.3 Factors

```{r}
fruit <- c("Apple", "Orange", "Banana")
parse_factor(c("Apple", "Orange", "Banana"),levels = fruit)
```

## 11.3.4 Dates, date-time and times

This will parse date-time according to ISO 8601

```{r}
parse_datetime(c("2018-09-02T1310"))
```

To parse time, we will use the hms package:

```{r}
library(hms)
parse_time("15:45")

```

## 11.3.5 Exercises

1. What are the most important arguments to locale()?

```{r}
?locale
```

From the readme: "A locale object tries to capture all the defaults that can vary between countries." Thus I would conclude
the most important arguments are the ones that specify the country.

2. What happens if you try and set decimal_mark and grouping_mark to the same character?
```{r}
#parse_character(x1,locale = locale(encoding = "Latin1",decimal_mark = ".", grouping_mark = "."))
```
<font color="red">Error: `decimal_mark` and `grouping_mark` must be different</font>



What happens to the default value of grouping_mark when you set decimal_mark to “,”? What happens to the default value of decimal_mark when you set the grouping_mark to “.”?

3. I didn’t discuss the date_format and time_format options to locale(). What do they do? Construct an example that shows when they might be useful.

```{r}
x1=parse_date(c("2018-09-02"))
x1
parse_character(x1,locale = locale(encoding = "Latin1"))
parse_character(x1, locale = locale(encoding = "Shift-JIS"))
```

5. What's the difference between read_csv and read_csv2?
read_csv2() uses ; for separators, instead of ,. This is common in European countries which use , as the decimal separator.

6. What are the most common encodings used in Europe? What are the most common encodings used in Asia? Do some googling to find out.
From Wikipedia:

    ISO 8859-1 Western Europe.
    ISO 8859-2 Western and Central Europe.
    ISO 8859-3 Western Europe and South European (Turkish, Maltese plus Esperanto)
    ISO 8859-4 Western Europe and Baltic countries (Lithuania, Estonia, Latvia and Lapp)
    ISO 8859-5 Cyrillic alphabet.
    ISO 8859-6 Arabic.
    ISO 8859-7 Greek.

Source: https://en.wikipedia.org/wiki/Character_encoding

7. Generate the correct format to parse each of the following correctly:

```{r}
d1 <- "January 1, 2010"

parse_datetime("January 1, 2010", "%B %d, %Y") # note MUST have comma in the format section for this to work

d2 <- "2015-Mar-07"
parse_datetime("2015-Mar-07", "%Y-%b-%d")

d3 <- "06-Jun-2017"
parse_datetime("06-jun-2017", "%d-%b-%Y")

d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14" # Dec 30, 2014
t1 <- "1705"
t2 <- "11:15:10.12 PM"
```


## 11.4.4 Parsing a file, (not just one line of a document)

From the text:

The heuristic tries each of the following types, stopping when it finds a match:

    logical: contains only “F”, “T”, “FALSE”, or “TRUE”.
    integer: contains only numeric characters (and -).
    double: contains only valid doubles (including numbers like 4.5e-5).
    number: contains valid doubles with the grouping mark inside.
    time: matches the default time_format.
    date: matches the default date_format.
    date-time: any ISO8601 date.

If none of these rules apply, then the column will stay as a vector of strings.

##11.4.3 Other Challenges:

```{r}
challenge2 <- read_csv(readr_example("challenge.csv"), guess_max = 1001)
challenge2
```

## 11.5, writing to a file

```{r}
write_csv(challenge2, "challenge.csv")
```

