---
title: "14 Strings.Rmd"
author: "Russ Conte"
date: "9/23/2018"
output: html_document
---

#14. Strings

## 14.1.1 Prerequisites

```{r}
library(tidyverse)
library(stringr)
```

##14.2 String Basics

Note - Hadley and Garrett recommend always using double quotes when quoting, not single quotes, unless quoting within a quote

```{r}
string1 <- "Four score and seven years ago"
string2 <- 'If I want to include a "quote" in a string, I\'ll use single quotes for the string and double quotes for the quote'
string1
string2
```

Note that the printed version of the string is *not* the same as the string itself. So see the string itself use writeLines(x)

```{r}
x <- c("\"", "\\")
x

writeLines(x) #interesting!
```

There are ways of writing non-English characters on all platforms. For example:

```{r}
x <- "\u00b5"
x #the Greek letter mu
```

Note that multiple strings are often stored as a character vector:

```{r}
c("First", "Second", "Third")
```

## 14.2.1 String Length

Note that we will be using stringr, and all functions start with str_.

```{r}
str_length(c("Now is the time", "for all good men", "to come to the aid of their country"))
```

## 14.2.2 Combining Strings

