---
title: "19 Functions"
author: "Russ Conte"
date: "10/21/2018"
output: html_document
---

# When should I write a function?

Whenever I've copied and pasted any block of code more than twice (and wow that has happened before!) Example to simplify:

```{r}
library(tidyverse)
df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c=rnorm(10),
  d=rnorm(10)
)
df$a <- (df$a - min(df$a, na.rm=TRUE))/(max(df$a, na.rm=TRUE - min(df$a, na.rm = TRUE)))
df$b <- (df$b - min(df$b, nb.rm=TRUE))/(max(df$b, nb.rm=TRUE - min(df$b, nb.rm = TRUE)))
df$c <- (df$c - min(df$c, nc.rm=TRUE))/(max(df$c, nc.rm=TRUE - min(df$c, nc.rm = TRUE)))
df$d <- (df$d - min(df$d, nd.rm=TRUE))/(max(df$d, nd.rm=TRUE - min(df$d, nd.rm = TRUE)))
df
df$a
df$b
df$c
df$d
```

Let's try a simpler version to get the same results:

```{r}
df$a
x <- df$a
rng <- range(x, na.rm=TRUE)
(x-rng[1]) / (rng[2]-rng[1])
```

```{r}
rescale01 <-  function(x){
  rng <-  range(x, na.rm=TRUE)
  (x-rng[1]) / (rng[2] - rng[1])
}

rescale01(c(0,5,10)) # note that the input is a vector, it did not work when I used (0,5,10)
rescale01(c(8,5,1,4,6,7,3,5))
rescale01(c(1,2,3,4,5,6,7,8))
```

Hadley and Garrett's three steps to creating a function:

1. Pick a good <i>name</i> for the function
2. List the inputs, also called arguments (and I prefer calling them inputs) inside the function
3. Place the code that I've developed in the body of the function
3.5 Check with some verifable examples, as well as edge cases, unique cases, impossible cases

```{r}
rescale01(c(-10,0,10))
rescale01(c(1,2,3,NA,5))
```

