---
title: "3. Data Transformation with dplyr"
author: "Russ Conte"
date: "8/5/2018"
output: html_document
---

##5.1.1 Prerequisites

These lines of code will setup our dataset, nycflights13, and tidyverse which will include dplyr.
```{r, echo=TRUE}
library(nycflights13)
library(tidyverse)
```

Let's have a look at the flights data set:

```{r}
flights
```

If you want to see the entire data set, use View(dataset), for example:

```{r}
View(flights)
```

## 5.1.3 dplyr basics: Select, Filter, Mutate, Arrange, Summarize (memorize these five terms!)


Pick observations by their values (filter()).
Reorder the rows (arrange()).
Pick variables by their names (select()).
Create new variables with functions of existing variables (mutate()).
Collapse many values down to a single summary (summarise()).
Note - these can <i>all</i> be used in conjunction with group_by()

How this works:

1. The first argument is the data frame
2. The second argument is the action (filter, arrange, etc)
3. The result is a new (not saved) data frame

Let's filter the flights to show flights on January 1, for all years:

```{r}
filter(flights, month==1, day==1)
```

Note that dplyr <i><b>never</i></b> saves the output, so you'll have to save it specifically if you want it later:

```{r}
jan1 <- filter(flights, day==1, day==1)
jan1
```

