```{r}
library(nycflights13)
library(tidyverse)
```

We will explore NYCflights13:

```{r}
nycflights13::airlines # a data set of airlines
```

```{r}
nycflights13::airports
```

```{r}
nycflights13::flights
```

```{r}
nycflights13::planes
```

```{r}
nycflights13::weather
```

```{r}
flights # provides a tibble of flights
View(flights)
```

From the text:

All verbs work similarly:

1. The first argument is a data frame.
2. The subsequent arguments describe what to do with the data frame, using the variable names (without quotes).
3. The result is a new data frame.

##### 5.2 Filter ROWS with filter() #####
 This will allow finding rows that match specific parameters:
 
```{r}
filter(flights, month == 1, day == 1) # finds all flights on January 1
```
 
 It is possible to save the result as a new data frame:
 
```{r}
feb02 <- filter(flights, month == 2, day == 2) # Now it's possible to do data analysis on this data set
feb02 #in R if a result is saved, it must be called to be viewed
```
 
 To print the results AND save them, enclose the command in parentheses:
 
```{r}
(nov1 <- filter(flights, month == 11, day == 1))
```
 
 Test the same idea with the Diamonds data set:
 
```{r}
diamonds
```
 
```{r}
(high_price_diamonds <- filter(diamonds, price>10000))
```
 
 5.2.2 Logical Operators: &, |, !
 
```{r}
filter(flights, month == 1 & arr_time>1000)
```
 
```{r}
filter(flights, month == 1 & arr_time>1000 & carrier == "B6")
```
 
 Here is a way to find all flights in November and December:
 
```{r}
nov_dec <- filter(flights, month == 11 | month == 12)
# An alternate way to do this search is to us %in%:

nov_december <- filter(flights, month %in% c(11,12))
nov_december
```
 
 5.2.3 Missing Values:
 To determine a missing value in R, use is.na()
 
```{r}
#Find all flights with missing departure times:
filter(flights, is.na(dep_time))
```
 
```{r}
# find all flights with missing sched_dep_time
filter(flights, is.na(sched_dep_time)) # none! all have data!
```
 
 5.2.4 Exercises
 
 1. Find all flights that had an arrival delay of two or more hours (let's do this several different ways)
 
```{r}
filter(flights, arr_delay>120)
```
 
```{r}
filter(flights, between(arr_delay, 120, 4800))
```
 
2. Flew to Houston (something I did myself in 2019)

```{r}
filter(flights, dest == "IAH" | dest == "HOU")
```

3. Were operated by United, American or Delta:

```{r}
filter(flights, carrier %in% c("UA", "DL", "AA"))
```

4. Departed in summer (July, August and September)

```{r}
filter(flights, month %in% c(7,8,9))
```

```{r}
filter(flights, between(month, 7, 9)) # same as above, different method
```

5. Arrived more than two hours late, but did not leave late:

```{r}
filter(flights,arr_delay>120 & dep_delay<=0)
```

6. Were delayed by at least an hour, but made up over 30 minutes in flight

```{r}
filter(flights, dep_delay>=60 & arr_delay <= 30)
```

7. Departed between midnight and 6:00 am (inclusive)

```{r}
filter(flights, between(dep_time, 0, 600))
```

3. How many flights have missing dep_time? 8,255

What other variables are missing? dep_delay, arr_time, arr_delay, air_time

What might these represent? Cancelled or rescheduled flights

```{r}
filter(flights, is.na(dep_time))
```

##### 5.3 Arrange rows with arrange() #####
From the text:
arrange() works similarly to filter() except that instead of selecting rows, it changes their order. It takes a data frame and a set of column names (or more complicated expressions) to order (the rows) by. If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns:

```{r}
arrange(flights, year, month, day)
```

```{r}
arrange(flights, desc(dep_delay))
```

```{r}
arrange(diamonds, desc(price))
```

```{r}
arrange(diamonds, desc(carat))
```

5.3.1 Exercises:

1. How could you use arrange() to sort all missing values to the start (hint: use is.na())

```{r}
arrange(flights, desc(is.na(dep_time)))
```

2. Sort flights to find the most delayed flights

```{r}
arrange(flights, desc(dep_delay))
```

Find the flights that left the earliest

```{r}
head(arrange(flights, dep_time), 200) # looking for the first 200 flights
```

3. Find the flights with the fastest (highest speed) flights

```{r}
speed <- mutate(flights, speed = distance/air_time) %>% 
  arrange(desc(speed))
speed
```

4. Which flights traveled the farthest?

```{r}
arrange(flights, desc(distance)) #farthest
```

Which flights traveled the shortest?
```{r}
arrange(flights,distance)
```

#### 5.4 Select Columns with Select ####

This returns a data frame with only those columns

```{r selecting three columns}
select(flights, year, month, day)
```

How to select columns between parameters:

```{r how to select columns between parameters}
select(flights, year:day)
```

How to select all columns EXCEPT the selected columns:

```{r selecting all columns EXCEPT the selected columns}
select(flights,-(year:day))
```

There are various select functions that are extremely useful, as follows:

starts_with("abc"): matches names that begin with “abc”.

```{r select using start with}
departure <- select(flights, starts_with("dep"))
departure
```

Let's find only the arrival data:

```{r selecting only the arrival columns}
select(flights, starts_with("arr"))
```

ends_with("xyz"): matches names that end with “xyz”.

```{r matching column names that end with specific characters or values}
select(flights, ends_with("time"))
```

contains("ijk"): matches names that contain “ijk”.

```{r select using Contains}
select(flights, contains("time"))
```

5.5 Add new variable with mutate()

```{r Mutate is a way to add new columns that are functions of existing columns}
flights_sml <- select(flights,
    year:day,
    ends_with("delay"),
    distance,
    air_time)
mutate(flights_sml,
       gain = dep_delay - arr_delay,
       speed = distance / air_time * 60)
```

It's possible to create new functions that refer to columns that were *just* created!

```{r Referring to columns that were just created}
mutate(flights_sml,
       gain = dep_delay - arr_delay,
       hours = air_time / 60,
       gain_per_hour = gain / hours)
```

#### 5.5.1 Useful Creation Functions ####

Modular arithmetic: %/% (integer division)
%% remainder

```{r Modular Arithmetic}
transmute(flights,
          dep_time,
          hour = dep_time %/% 100,
          minute = dep_time %% 100)
```


Offsets, lead() and lag() allow me to refer to leading or lagging values, and calculate running differences (x - lag(x))

```{r leading and lagging values}
(x <- rnorm(1:60))
lag(x)
lead(x)
```

Cumulative and rolling aggregates:

```{r cumulative and rolling aggregates}
cumsum(x)
cummean(x)
```


#### 5.6 Grouped summaries with summarise()

<h1>Together group_by() and summarise() provide one of the tools that you’ll use most commonly when working with dplyr: grouped summaries.</h1>

```{r summarise example gives average delay by date}
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

#### 5.6.1 Combining Multiple Operations with the Pipe %>% 

%>% ~ "then"
Example: "we want to explore the relationship between the distance and average delay for each location."

```{r first example using the pipe starting with an example that does not use the pipe command}
by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE)
)
delay <- filter(delay, count > 20, dest != "HNL")

# It looks like delays increase with distance up to ~750 miles 
# and then decrease. Maybe as flights get longer there's more 
# ability to make up delays in the air?
ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)

```


There are three steps to prepare this data:

Group flights by destination.

Summarise to compute distance, average delay, and number of flights.

Filter to remove noisy points and Honolulu airport, which is almost twice as far away as the next closest airport.

```{r now we will do exactly the same analysis using the pipe command and note the line to drop groups}
library(tidyverse)
delays <- flights %>%
  group_by(dest) %>%
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean (arr_delay, na.rm = TRUE),
    .groups = 'drop') %>% # https://stackoverflow.com/questions/62140483/how-to-interpret-dplyr-message-summarise-regrouping-output-by-x-override
    filter(count>20, dest != "HNL")

ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)

```

#### 5.6.2 Missing Values
Example: What happens if we don't set na.rm?

```{r not seeting na.rm what happens - we get a lot of missing values}
flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))


```

Now let's do the same, but remove missing values:

```{r removing missing values}
flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay, na.rm = TRUE)) %>% 
  ggplot(mapping = aes(x = month)) +
  geom_bar()

# note best month = February

```

Next we remove the cancelled flights, assuming missing values = cancelled flights:

```{r removing cancelled flights}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))
not_cancelled


```

#### 5.6.3 Counts ####

let's look at planes (identified by tail number) that have the highest average delays:

```{r find the planes that have the highest average delays}

delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay)
  )

ggplot(data = delays, mapping = aes(x = delay)) +
  geom_freqpoly(binwidth = 10)
```

```{r same as above but with bargraph}
delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay)
  )

ggplot(data = delays, mapping = aes(x = delay)) +
  geom_histogram(binwidth = 1)


```

sort planes by delay, biggest to smallest:

```{r sort planes by delay, biggest to smallest}
arrange(delays, desc(delay))

```

The story is a bit more nuanced - draw a scatter plot of number of flights vs delay:

```{r scatterplot of number of flights vs delay}

delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )
ggplot(data = delays, mapping = aes(x = n, y = delay)) +
geom_point(alpha = 1/10)

```

From the text:
<h1>The shape of this plot is very characteristic: whenever you plot a mean (or other summary) vs. group size, you’ll see that the variation decreases as the sample size increases.</h1>

```{r combining ggplot and dplyr into one workflow}
delays %>% 
  filter(n>25) %>% 
  ggplot(mapping = aes(x = n, y = delay)) +
    geom_point(alpha = 1/10)

```

from the text: " Let’s look at how the average performance of batters in baseball is related to the number of times they’re at bat."

"When I plot the skill of the batter (measured by the batting average, ba) against the number of opportunities to hit the ball (measured by at bat, ab), you see two patterns:

As above, the variation in our aggregate decreases as we get more data points.

There’s a positive correlation between skill (ba) and opportunities to hit the ball (ab). This is because teams control who gets to play, and obviously they’ll pick their best players."

```{r plotting at bats vs batting average}
batting <- as_tibble(Lahman::Batting)

batters <- batting %>% 
  group_by(playerID) %>% 
  summarise(
    ba = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    ab = sum(AB, na.rm = TRUE))

batters %>% 
  filter(ab>100) %>% 
  ggplot(mapping = aes(x = ab, y = ba)) +
  geom_point() +
  geom_smooth(se = FALSE)

```

#### Useful summary functions ####

```{r an example of subsetting finding average delay and average positive delay}

not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    avg_delay1 = mean(arr_delay),
    avg_delay2 = mean(arr_delay[arr_delay>0]),
    avg_delay3 = mean(arr_delay[arr_delay<=0])
  )


```

Counts - some good examples from Haley :)

```{r counting the number of non-missing values}
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(carriers = n_distinct(carrier)) %>% 
  arrange(desc(carriers))

```

dplyr includes a simple function if all I need is a count:

```{r a dplyr function if all I need is a count, so count number of flights by dest}
not_cancelled %>% 
  count(dest)


```

Let's try the same thing with the Diamonds data set

```{r counting in the Diamonds dataset}

diamonds %>% 
  count(color)

```

```{r Count number of diamonds by Cut}
diamonds %>% 
  count(cut)
```

```{r an example of counting by TeamID using the built-in dplyr function}
Lahman::Batting
batting %>% 
  count(teamID)

```

#### 5.6.5 Grouping by multiple variables ####

From the text: When you group by multiple variables, each summary peels off one level of the grouping. That makes it easy to progressively roll up a dataset:

```{r grouping by multiple variables}
daily <- group_by(flights, year, month, day)
(per_day <- summarise(daily, flights = n())) %>% 
  ggplot(mapping = aes(x = flights)) +
  geom_bar()

```

