---
title: "6 Scripts"
author: "Russ Conte"
date: "8/19/2018"
output: html_document
---

# 6. Scripts

## 6.1 Running Code

#from the text: "The key to using the script editor effectively is to memorise one of the most important keyboard shortcuts: Cmd/Ctrl + Enter.
#This executes the current R expression in the console."

#If your cursor is at █, pressing Cmd/Ctrl + Enter will run the complete command that generates not_cancelled.
#It will also move the cursor to the next statement (beginning with not_cancelled %>%).
#That makes it easy to run your complete script by repeatedly pressing Cmd/Ctrl + Enter.



```{r setup, include=FALSE}
library(dplyr)
library(nycflights13)

not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))
```

There is a way to run the entire script, not just line-by-line: Cmd/Ctrl + Shift + Enter for the current chunk,
and Option + Command + R to run everything.

```{r setup, include=FALSE}
library(dplyr)
library(nycflights13)

not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))
```
