---
title: "7 Exploratory Data Analysis"
author: "Russ Conte"
date: "8/19/2018"
output: html_document
---

Hadley and Garrett state that in Exploratory Data Analysis (EDA) you:
      Generate questions about your data.
      Search for answers by visualising, transforming, and modelling your data.
      Use what you learn to refine your questions and/or generate new questions.

##7.1.1 Prerequisites for EDA:

```{r}
library(tidyverse)
```

From the text: Your goal during EDA is to develop an understanding of your data. The easiest way to do this is to <i><b>use questions as tools to guide your investigation</i></b>. When you ask a question, the question focuses your attention on a specific part of your dataset and helps you decide which graphs, models, or transformations to make.

Two types of questions will always be useful for making discoveries within your data. You can loosely word these questions as:

    What type of <i>variation</i> occurs <i>within</i> my variables?

    What type of <i>covariation<i/> occurs <i>between</i> my variables?

##7.3 Variation

From the text (this is another fantastic truth in the text!): "The best way to understand that pattern is to visualise the distribution of the variable’s values."

##7.3.1 Visualizing distributions

An example of the distribution of a categorical variable:

```{r}
ggplot(data=diamonds) +
  geom_bar(mapping=aes(x=cut))
```

Note the height of the bars = count by type of cut. These values can be shown directly:

```{r}
diamonds %>% 
  count(cut)
```

One way to look at <i>continuous</i> values is via a histogram:

```{r}
diamonds
ggplot(data=diamonds) +
  geom_histogram(mapping = aes(x=price),binwidth = 10)
```

Interesting idea: Let's look at count of diamonds with carat<3:

```{r}
smaller <- filter(diamonds, carat<3)
bigger <- filter(diamonds, carat>=3)

ggplot(data=smaller)+
  geom_histogram(mapping = aes(x=carat), binwidth = 0.01)

ggplot(data=bigger)+
  geom_histogram(mapping = aes(x=carat), binwidth = 0.001)
```

One way to put multiple plots on the same graph is to use freqploy:

```{r}
smaller
ggplot(data=smaller, mapping=aes(x=carat, color=cut))+
  geom_freqpoly(binwidth=0.1)

ggplot(data=smaller, mapping=aes(x=carat, color=color))+
  geom_freqpoly(binwidth=0.1)

ggplot(data=smaller, mapping=aes(x=carat, color=price))+
  geom_freqpoly(binwidth=0.1)
```

Another great line from Hadley and Garrett:

"The key to asking good follow-up questions will be to rely on your curiosity (What do you want to learn more about?) as well as your skepticism (How could this be misleading?)."

##7.3.2 Typical Values

As an example of typical questions, look at this histogram:

```{r}
ggplot(data=smaller) +
  geom_histogram(mapping = aes(x=carat),binwidth = 0.01)
```

The authors ask a few good questions, here are some answers:

Why are there more diamonds at whole carats and common fractions of carats?
I don't know, it is a testable question that diamonds in this set were cut to those specifications, but that is a hypothesis not a fact

Why are there more diamonds slightly to the right of each peak than there are slightly to the left of each peak?
Possibly because those diamonds sell for a higher cost than diamonds slightly below whole value of carats. That is a hypothesis not a fact

Why are there no diamonds bigger than 3 carats?
Because the data set removed all diamonds >3 carats.

Let's practice wit data of eruptions of Old Faithful in Yellowstone:

```{r}
faithful
ggplot(data=faithful) +
  geom_histogram(mapping = aes(x=eruptions),binwidth = 0.25)
```

##7.3.3 Unusual values (outliers)

Notice there are no visible values >10, but the x-axis extends out to 60. There are some values >10, but not many. Let's investigate!

```{r}
ggplot(diamonds) +
  geom_histogram(mapping = aes(x=y), binwidth = 0.5)
```

```{r}
ggplot(diamonds)+
  geom_histogram(mapping = aes(x=y), binwidth = 0.5)+
  coord_cartesian(ylim=c(0,50))
```

Let's use dplyr to find the three values that are outliers:

```{r}
unusual <- diamonds %>% 
  filter(y<3 | y>20) %>% 
  select(price, x,y,z) %>% 
  arrange(y)
unusual
```

More great advice from Hadley and Garrett:

"It’s good practice to repeat your analysis with and without the outliers. If they have minimal effect on the results, and you can’t figure out why they’re there, it’s reasonable to replace them with missing values, and move on. However, if they have a substantial effect on your results, you shouldn’t drop them without justification. You’ll need to figure out what caused them (e.g. a data entry error) and disclose that you removed them in your write-up."

##7.3.4 Exercises

Explore the distribution of each of the x, y, and z variables in diamonds. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.

```{r}
ggplot(data=diamonds)+
  geom_histogram(mapping = aes(x=x),binwidth = 0.25)

ggplot(data=diamonds)+
  geom_histogram(mapping = aes(x=y),binwidth = 0.25)

ggplot(data=diamonds)+
  geom_histogram(mapping = aes(x=z),binwidth = 0.25)

```

Note: The x and y values range from 0 through 60, z values range from 0 to approximately 30.

2. Explore the distribution of price. Do you discover anything unusual or surprising? (Hint: Carefully think about the binwidth and make sure you try a wide range of values.)

```{r}
ggplot(data=diamonds) +
  geom_histogram(mapping = aes(x=price), binwidth = 1)

ggplot(data=diamonds) +
  geom_histogram(mapping = aes(x=price), binwidth = 10)

ggplot(data=diamonds) +
  geom_histogram(mapping = aes(x=price), binwidth = 100)

ggplot(data=diamonds) +
  geom_histogram(mapping = aes(x=price), binwidth = 1000)

ggplot(data=diamonds) +
  geom_histogram(mapping = aes(x=price), binwidth = 10000)


```

The number of diamonds goes down as price goes up. The vast majority of diamonds are <$10,000, a few are >$15,000

3. How many diamonds are 0.99 carats? How many are 1 carat? What do you think accounts for this discrepancy?

```{r}
diamonds %>% 
  count(carat==0.99) #23 diamonds are 0.99 carat

diamonds %>% 
  count(carat==1.00) #1558 diamonds are 1.00 carat

```

My hypothesis is that diamonds that are 1.0 carat sell easier than diamonds that are 0.99 carats.

4. Compare and contrast coord_cartesian() vs xlim() or ylim() when zooming in on a histogram. What happens if you leave binwidth unset? What happens if you try and zoom so only half a bar shows?
```{r}
ggplot(diamonds)+
  geom_histogram(mapping = aes(x=y), binwidth = 0.5)+
  coord_cartesian(ylim=c(0,12000))

ggplot(diamonds)+
  geom_histogram(mapping = aes(x=y), binwidth = 0.5) +
  xlim(0,100)

ggplot(diamonds)+
  geom_histogram(mapping = aes(x=y)) +
  xlim(0,100)
```

## 7.4 Missing Values

One way to analyze unusual values is the replace them with NA, here we replace y<3 or y>20 with NA:

```{r}
diamonds2 <- diamonds %>% 
  mutate(y=ifelse(y<3 | y>20, NA, y))
```

Note that ggplot (like R) will report on missing values:

```{r}
ggplot(data=diamonds2,mapping = aes(x=x, y=y)) +
  geom_point()
```

If we want to supress that warning, use na.rm=TRUE:

```{r}
ggplot(data=diamonds2, mapping = aes(x=x, y=y)) +
  geom_point(na.rm = TRUE)
```

Let's look at what makes missing values different from observed values, using the NYCFlights13 data set:

```{r}
nycflights13::flights %>% 
  mutate(
    cancelled=is.na(dep_time),
    sched_hour=sched_dep_time%/%100,
    sched_min=sched_dep_time%%100,
    sched_dep_time=sched_hour+sched_min/60
    ) %>% 
  ggplot(mapping = aes(sched_dep_time)) +
           geom_freqpoly(mapping = aes(color=cancelled), binwidth=0.25)
```

The authors say (and I totally agree) this plot does not tell us a lot, mainly because there are few missing flights compared to
scheduled flights. They write that they will address in a future section of the book.

## 7.4.1 Exercises

1. What happens to missing values in a histogram? Barchart? Why is there a difference?

```{r}
miss=data.frame(a1=c(1:10,NA,NA), b=c(1:12))
miss
ggplot(miss) +
  geom_histogram(mapping = aes(x=a1)) #histogram removes the rows with missing data

ggplot(miss) +
  geom_bar(mapping = aes(x=a1)) # bar also removes one row containing missing data

```

2. What does na.rm = TRUE do in mean() and sum()?

```{r}
miss
mean(miss$a1) # with NA it returns NA

mean(miss$a1,na.rm = TRUE) #it returns the value (5.5)

sum(miss$a1) #returns NA

sum(miss$a1, na.rm=TRUE) #returns the sum of 55
```

## 7.5 Covariation

Covariation describes variables that vary together (such as height and weight in humans). It's not always easy to see:

```{r}
ggplot(diamonds, mapping = aes(x=price)) +
  geom_freqpoly(mapping = aes(color=cut), binwidth=500)
```

The differences are difficult to see because the counts differ so much:

```{r}
ggplot(diamonds,aes(x=cut)) +
  geom_bar()
```

What we'll do is standardize the values, so the area under each curve=1.

```{r}
ggplot(diamonds, mapping = aes(x=price, y=..density..)) +
  geom_freqpoly(mapping = aes(color=cut), binwidth=500)
```

let's look at another way to visualize the data - a boxplot.

```{r}
ggplot(diamonds, aes(x=cut, y=price)) +
  geom_boxplot(mapping = aes(color=cut))
```

Another example is to look at how highway mileage varies across class of car:

```{r}
ggplot(mpg, aes(x=class, y=hwy)) +
  geom_boxplot(mapping = aes(color=class))
```

Let's reorder the data to make it easier to understand:

```{r}
ggplot(mpg) +
  geom_boxplot(mapping = aes(x=reorder(class, hwy, FUN=mean), y=hwy))
```

## 7.5.1 Exercises