---
title: "7 Exploratory Data Analysis"
author: "Russ Conte"
date: "8/19/2018"
output: html_document
---

Hadley and Garrett state that in Exploratory Data Analysis (EDA) you:
      Generate questions about your data.
      Search for answers by visualising, transforming, and modelling your data.
      Use what you learn to refine your questions and/or generate new questions.

##7.1.1 Prerequisites for EDA:

```{r}
library(tidyverse)
```

From the text: Your goal during EDA is to develop an understanding of your data. The easiest way to do this is to <i><b>use questions as tools to guide your investigation</i></b>. When you ask a question, the question focuses your attention on a specific part of your dataset and helps you decide which graphs, models, or transformations to make.

Two types of questions will always be useful for making discoveries within your data. You can loosely word these questions as:

    What type of <i>variation</i> occurs <i>within</i> my variables?

    What type of <i>covariation<i/> occurs <i>between</i> my variables?

##7.3 Variation

From the text (this is another fantastic truth in the text!): "The best way to understand that pattern is to visualise the distribution of the variable’s values."

##7.3.1 Visualizing distributions

An example of the distribution of a categorical variable:

```{r}
ggplot(data=diamonds) +
  geom_bar(mapping=aes(x=cut))
```

Note the height of the bars = count by type of cut. These values can be shown directly:

```{r}
diamonds %>% 
  count(cut)
```

One way to look at <i>continuous</i> values is via a histogram:

```{r}
diamonds
ggplot(data=diamonds) +
  geom_histogram(mapping = aes(x=price),binwidth = 10)
```

Interesting idea: Let's look at count of diamonds with carat<3:

```{r}
smaller <- filter(diamonds, carat<3)
bigger <- filter(diamonds, carat>=3)

ggplot(data=smaller)+
  geom_histogram(mapping = aes(x=carat), binwidth = 0.01)

ggplot(data=bigger)+
  geom_histogram(mapping = aes(x=carat), binwidth = 0.001)
```

One way to put multiple plots on the same graph is to use freqploy:

```{r}
smaller
ggplot(data=smaller, mapping=aes(x=carat, color=cut))+
  geom_freqpoly(binwidth=0.1)

ggplot(data=smaller, mapping=aes(x=carat, color=color))+
  geom_freqpoly(binwidth=0.1)

ggplot(data=smaller, mapping=aes(x=carat, color=price))+
  geom_freqpoly(binwidth=0.1)
```

Another great line from Hadley and Garrett:

"The key to asking good follow-up questions will be to rely on your curiosity (What do you want to learn more about?) as well as your skepticism (How could this be misleading?)."

##7.3.2 Typical Values

As an example of typical questions, look at this histogram:

```{r}
ggplot(data=smaller) +
  geom_histogram(mapping = aes(x=carat),binwidth = 0.01)
```

The authors ask a few good questions, here are some answers:

Why are there more diamonds at whole carats and common fractions of carats?
I don't know, it is a testable question that diamonds in this set were cut to those specifications, but that is a hypothesis not a fact

Why are there more diamonds slightly to the right of each peak than there are slightly to the left of each peak?
Possibly because those diamonds sell for a higher cost than diamonds slightly below whole value of carats. That is a hypothesis not a fact

Why are there no diamonds bigger than 3 carats?
Because the data set removed all diamonds >3 carats.

Let's practice wit data of eruptions of Old Faithful in Yellowstone:

```{r}
faithful
ggplot(data=faithful) +
  geom_histogram(mapping = aes(x=eruptions),binwidth = 0.25)
```

##7.3.3 Unusual values (outliers)

Notice there are no visible values >10, but the x-axis extends out to 60. There are some values >10, but not many. Let's investigate!

```{r}
ggplot(diamonds) +
  geom_histogram(mapping = aes(x=y), binwidth = 0.5)
```

```{r}
ggplot(diamonds)+
  geom_histogram(mapping = aes(x=y), binwidth = 0.5)+
  coord_cartesian(ylim=c(0,50))
```

Let's use dplyr to find the three values that are outliers:

```{r}
unusual <- diamonds %>% 
  filter(y<3 | y>20) %>% 
  select(price, x,y,z) %>% 
  arrange(y)
unusual
```

More great advice from Hadley and Garrett:

"It’s good practice to repeat your analysis with and without the outliers. If they have minimal effect on the results, and you can’t figure out why they’re there, it’s reasonable to replace them with missing values, and move on. However, if they have a substantial effect on your results, you shouldn’t drop them without justification. You’ll need to figure out what caused them (e.g. a data entry error) and disclose that you removed them in your write-up."

##7.3.4 Exercises

Explore the distribution of each of the x, y, and z variables in diamonds. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.

```{r}
ggplot(data=diamonds)+
  geom_histogram(mapping = aes(x=x),binwidth = 0.25)

ggplot(data=diamonds)+
  geom_histogram(mapping = aes(x=y),binwidth = 0.25)

ggplot(data=diamonds)+
  geom_histogram(mapping = aes(x=z),binwidth = 0.25)

```

Note: The x and y values range from 0 through 60, z values range from 0 to approximately 30.

2. Explore the distribution of price. Do you discover anything unusual or surprising? (Hint: Carefully think about the binwidth and make sure you try a wide range of values.)

```{r}
ggplot(data=diamonds) +
  geom_histogram(mapping = aes(x=price), binwidth = 1)

ggplot(data=diamonds) +
  geom_histogram(mapping = aes(x=price), binwidth = 10)

ggplot(data=diamonds) +
  geom_histogram(mapping = aes(x=price), binwidth = 100)

ggplot(data=diamonds) +
  geom_histogram(mapping = aes(x=price), binwidth = 1000)

ggplot(data=diamonds) +
  geom_histogram(mapping = aes(x=price), binwidth = 10000)


```

The number of diamonds goes down as price goes up. The vast majority of diamonds are <$10,000, a few are >$15,000

3. How many diamonds are 0.99 carats? How many are 1 carat? What do you think accounts for this discrepancy?

```{r}
diamonds %>% 
  count(carat==0.99) #23 diamonds are 0.99 carat

diamonds %>% 
  count(carat==1.00) #1558 diamonds are 1.00 carat

```

My hypothesis is that diamonds that are 1.0 carat sell easier than diamonds that are 0.99 carats.

4. Compare and contrast coord_cartesian() vs xlim() or ylim() when zooming in on a histogram. What happens if you leave binwidth unset? What happens if you try and zoom so only half a bar shows?
```{r}
ggplot(diamonds)+
  geom_histogram(mapping = aes(x=y), binwidth = 0.5)+
  coord_cartesian(ylim=c(0,12000))

ggplot(diamonds)+
  geom_histogram(mapping = aes(x=y), binwidth = 0.5) +
  xlim(0,100)

ggplot(diamonds)+
  geom_histogram(mapping = aes(x=y)) +
  xlim(0,100)
```

## 7.4 Missing Values

One way to analyze unusual values is the replace them with NA, here we replace y<3 or y>20 with NA:

```{r}
diamonds2 <- diamonds %>% 
  mutate(y=ifelse(y<3 | y>20, NA, y))
```

Note that ggplot (like R) will report on missing values:

```{r}
ggplot(data=diamonds2,mapping = aes(x=x, y=y)) +
  geom_point()
```

If we want to supress that warning, use na.rm=TRUE:

```{r}
ggplot(data=diamonds2, mapping = aes(x=x, y=y)) +
  geom_point(na.rm = TRUE)
```

Let's look at what makes missing values different from observed values, using the NYCFlights13 data set:

```{r}
nycflights13::flights %>% 
  mutate(
    cancelled=is.na(dep_time),
    sched_hour=sched_dep_time%/%100,
    sched_min=sched_dep_time%%100,
    sched_dep_time=sched_hour+sched_min/60
    ) %>% 
  ggplot(mapping = aes(sched_dep_time)) +
           geom_freqpoly(mapping = aes(color=cancelled), binwidth=0.25)
```

The authors say (and I totally agree) this plot does not tell us a lot, mainly because there are few missing flights compared to
scheduled flights. They write that they will address in a future section of the book.

## 7.4.1 Exercises

1. What happens to missing values in a histogram? Barchart? Why is there a difference?

```{r}
miss=data.frame(a1=c(1:10,NA,NA), b=c(1:12))
miss
ggplot(miss) +
  geom_histogram(mapping = aes(x=a1)) #histogram removes the rows with missing data

ggplot(miss) +
  geom_bar(mapping = aes(x=a1)) # bar also removes one row containing missing data

```

2. What does na.rm = TRUE do in mean() and sum()?

```{r}
miss
mean(miss$a1) # with NA it returns NA

mean(miss$a1,na.rm = TRUE) #it returns the value (5.5)

sum(miss$a1) #returns NA

sum(miss$a1, na.rm=TRUE) #returns the sum of 55
```

## 7.5 Covariation

Covariation describes variables that vary together (such as height and weight in humans). It's not always easy to see:

```{r}
ggplot(diamonds, mapping = aes(x=price)) +
  geom_freqpoly(mapping = aes(color=cut), binwidth=500)
```

The differences are difficult to see because the counts differ so much:

```{r}
ggplot(diamonds,aes(x=cut)) +
  geom_bar()
```

What we'll do is standardize the values, so the area under each curve=1.

```{r}
ggplot(diamonds, mapping = aes(x=price, y=..density..)) +
  geom_freqpoly(mapping = aes(color=cut), binwidth=500)
```

let's look at another way to visualize the data - a boxplot.

```{r}
ggplot(diamonds, aes(x=cut, y=price)) +
  geom_boxplot(mapping = aes(color=cut))
```

Another example is to look at how highway mileage varies across class of car:

```{r}
ggplot(mpg, aes(x=class, y=hwy)) +
  geom_boxplot(mapping = aes(color=class))
```

Let's reorder the data to make it easier to understand:

```{r}
ggplot(mpg) +
  geom_boxplot(mapping = aes(x=reorder(class, hwy, FUN=mean), y=hwy))
```

## 7.5.1 Exercises

1. Use what you’ve learned to improve the visualisation of the departure times of cancelled vs. non-cancelled flights.

```{r}
library(nycflights13)
nycflights13::flights %>% 
  mutate(
    cancelled=is.na(dep_time),
    sched_hour=sched_dep_time%/%100,
    sched_min=sched_dep_time%%100,
    sched_dep_time=sched_hour+sched_min/60
    ) %>% 
  ggplot(mapping = aes(x=cancelled, y=sched_dep_time)) +
             geom_boxplot()
```

** note** the on-time flights have an ealier departure time EXCEPT for one outlier under TRUE that is close to 0

2. What variable in the diamonds dataset is most important for predicting the price of a diamond? How is that variable correlated with cut? Why does the combination of those two relationships lead to lower quality diamonds being more expensive?

```{r}
diamonds.lm=lm(price~., data=diamonds)
summary(diamonds.lm) #Carat has highest beta with price (11,256.978)

#How is that variable correlated with cut?
cor(diamonds$price, diamonds$carat) # the correlation is 0.9215913, quite high

#Why does the combination of those two relationships lead to lower quality diamonds being more expensive?
#Because all of the quality variables in the linear model have a strongly negative coefficient
```

3. Install the ggstance package, and create a horizontal boxplot. How does this compare to using coord_flip()?

```{r}
library(ggstance)
library(ggplot2)

# using ggplot2:
ggplot(diamonds, aes(price, carat, fill=factor(cut))) +
  geom_boxplot() +
  coord_flip()

#using ggstance:
ggplot(diamonds, aes(price, carat, fill = factor(cut))) +
  geom_boxploth()
```

Note that the prior boxplots have a lot of outlier points!

From the text: "One problem with boxplots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of “outlying values”. One approach to remedy this problem is the letter value plot. Install the lvplot package, and try using geom_lv() to display the distribution of price vs cut. What do you learn? How do you interpret the plots?"

```{r}
diamonds
library(ggplot2)
library(lvplot)
p <- ggplot(diamonds, aes(x=cut, y=price))
p + geom_lv(aes(fill=..LV..))
```

These plots are stacked, unlike the former boxplots, and decrease in size as the price goes up.

5. Compare and contrast geom_violin() with a facetted geom_histogram(), or a coloured geom_freqpoly(). What are the pros and cons of each method?

```{r}
ggplot(diamonds, aes(price, carat, fill=factor(cut))) +
  geom_violin()
```

```{r}
diamonds
library(ggplot2)
ggplot(diamonds, aes(x=price)) +
  geom_histogram() +
  facet_wrap(~color)

ggplot(diamonds, aes(x=price)) +
  geom_freqpoly() +
  facet_wrap(~color)

```

I find the faceted histogram to be the easiest to understand and interpret.

Let's look a ggbeeswarm:

```{r}
diamonds
library(tidyverse)
library(ggbeeswarm)
ggplot(diamonds, aes(x=cut, y=price)) +
  geom_beeswarm()
```

## 7.5.2 Two categorical variables

We can visualize covariation between two categorical variables. One way is to use the built in count function:

```{r}
ggplot(data=diamonds) +
  geom_count(mapping = aes(x=cut, y=color))
```

We can also use dplyr to calculate the count:

```{r}
library(tidyverse)
diamonds %>%
  count(cut,color)
```

Now we can visualize the results with ggplot :)

```{r}
diamonds
library(tidyverse)
diamonds %>%
  count(cut,color) %>% 
  ggplot(mapping=aes(x=cut, y=color)) +
  geom_tile(mapping = aes(fill=n))
```

## 7.5.3 Two Continuous Variables

One way to see the relationship between two continuous variables is with a scatterplot:

```{r}
library(tidyverse)
ggplot(diamonds, aes(x=carat, y=price)) +
  geom_point()
```

Let's see how this looks using the package 'hexbin':

```{r}
library(tidyverse)
ggplot(smaller) +
  geom_bin2d(aes(x=carat, y=price))
```

```{r}
library(hexbin)
ggplot(smaller) +
  geom_hex(aes(x=carat, y=price))
```

Another very interesting example is to bin a continuous variable (such as carat), and then use the tools
to plot categorical variables. The first soluation puts approximately the same value of carate, the width is independent of the number of observations (note 'cut_width'):

```{r}
ggplot(smaller, aes(x=carat, y=price)) +
  geom_boxplot(aes(group=cut_width(carat, 0.1)))
```

To plot depending on the number of observations (note 'cut_number'):

```{r}
library(tidyverse)
ggplot(smaller, aes(x=carat, y=price)) +
  geom_boxplot(aes(group=cut_number(carat, 20)))
```

## 7.5.3.1 Exercises:

1. Instead of summarising the conditional distribution with a boxplot, you could use a frequency polygon. What do you need to consider when using cut_width() vs cut_number()? How does that impact a visualisation of the 2d distribution of carat and price?

cut_number puts the same number of observations into each bin, cut_width makes each bin the same size.

2. Visualize carat, partitioned by price.

```{r}
library(tidyverse)
ggplot(diamonds, aes(x=carat, y=price)) +
  geom_boxplot(aes(group=cut_number(price, 10)))
```

3. How does the price distribution of very large diamonds compare to small diamonds? Is it as you expect, or does it surprise you?

```{r}
very_large <- filter(diamonds, carat>2)
small <- filter(diamonds, carat<1)
library(tidyverse)
ggplot(very_large, aes(x=carat, y=price)) +
  geom_boxplot()

ggplot(small, aes(x=carat, y=price)) +
  geom_boxplot(aes(group=cut_number(price, 10)))

```

Analysis: There are fewer large diamonds than small diamonds. The price distribution of very large diamonds is
much narrower than small diamonds. The few large diamonds are in one boxplot, but the many more smaller diamonds
are in 10 boxplots.

4. Combine two of the techniques you’ve learned to visualise the combined distribution of cut, carat, and price.

```{r}
library(tidyverse)
ggplot(diamonds, aes(x=cut, y=carat)) +
  geom_boxplot(aes(group=cut_number(price, 10)))
```

5. Why is a scatterplot better than a binned plot for this case: 
  ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y)) +
  coord_cartesian(xlim = c(4, 11), ylim = c(4, 11))
  
```{r}
  ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y)) +
  coord_cartesian(xlim = c(4, 11), ylim = c(4, 11))
```

Because the binned plots would obscure some of the points

## 7.6 Patterns and Models

From the text: "Once you’ve removed the strong relationship between carat and price, you can see what you expect in the relationship between cut and price: relative to their size, better quality diamonds are more expensive."

```{r}
library(tidyverse)
library(modelr)
mod <-lm(log(price)~log(carat), data=diamonds)

diamonds2 <-  diamonds %>% 
  add_residuals(mod) %>% 
  mutate(resid=exp(resid))

ggplot(diamonds2) +
  geom_point(aes(x=carat, y=resid))


```

From the text: "Once you’ve removed the strong relationship between carat and price, you can see what you expect in the relationship between cut and price: relative to their size, better quality diamonds are more expensive."

```{r}
ggplot(diamonds2) +
  geom_boxplot(aes(x=cut, y=resid))
```

## 7.7 gggplot2 calls

It's possible to condense a lot of the data in ggplot 2 calls:

```{r}
ggplot(data=faithful, mapping=aes(x=eruptions)) +
  geom_freqpoly(binwidth=0.25)  
```

is the same as (note the condensed code):

```{r}
ggplot(faithful, aes(x=eruptions)) +
  geom_freqpoly(binwidth=0.25)
```

This is a test.