---
title: "9. Tibbles.Rmd"
author: "Russ Conte"
date: "9/1/2018"
output: html_document
---

Introduction to tibbles

1. How to make tibbles:

```{r}
library(tidyverse)
iris.tibble <- as_tibble(iris)
class(iris.tibble)

tibble(
  x=1:15,
  y=1,
  z=x^2+y
)

#unusual names for tibbles:

tb <- tibble(
  ':)' = "smile",
  ' ' = "space",
  '2000' = "number"
)

tb

```

How to use tribble() (transposed tibble) to make a tibble - note this transposes columns and rows, and it is
customized for data entry in code:

```{r}
tribble(
  ~x, ~y, ~z,
  #--/--/----
  "a",2,3.6,
  "b",1,8.5
)
```

Tibbles Versus data.frame: Two big differences: Printing and subsetting

```{r}
tibble(
  a=lubridate::now() + runif(1e3) * 86400,
  b=lubridate::today() + runif(1e3) * 30,
  c=1:1e3,
  d=runif(1e3),
  e=sample(letters, 1e3, replace=TRUE)
)
```

How to print tibbles:

```{r}
nycflights13::flights %>% 
  print(n=10, width=inf)
```

## Subsetting tibbles

```{r}
df <- tibble(
  x=runif(5),
  y=rnorm(5)
)
df

#extract by name:

df$x

df$y

#extract by position:

df[[1]]

df[[1]][2]

#to use these in a pipe, it requires a special placeholder, the period (or dot):

df %>% .$x

df %>% .[["x"]]

class(df)

```


## Exercises

1. How can you tell if an object is a tibble? (hint - try printing mtcars, which is a regular data frame)
One way is class(df)
```{r}
print(mtcars)
mtcars.tibble <- as.tibble(mtcars)
class(mtcars.tibble)
print(mtcars.tibble)

```

2. Compare and contrast the following operations on a data frame and an equivalent tibble. What is the difference?

```{r}
df <- data.frame(abc=1, xyz="a")
df$x
df[,"xyz"]
df[,c("abc", "xyz")]


# 3. If you have the name of a variable stored in an object, e.g. var <- "mpg", how can you extract the *reference* variable from the tibble?
tibble1 <- tibble(abc=1, xyz="a")
tibble1[[1]]

```

```{r}
annoying <- tibble(
  `1`=1:10,
  `2` = `1`*2 + rnorm(length(`1`))
)

annoying #look at the data set

# extract the variable called `1`

annoying$`1`

#print a scatterplot of 1 versus 2
ggplot(annoying, aes(x=`1`, y=`2`)) +
  geom_point()

# create a new column which is 2 / 1

annoying$`3` <- annoying$`2`/annoying$`1`

#rename the columns to one, two, three
annoying %>% 
  rename(one = `1`, two=`2`, three = `3`)

```

