---
title: "3 Data Visualization Example"
author: "Russ Conte"
date: "8/4/2018"
output: html_document
---
This chapter will be a look at ggplot2. ggplot2 is part of tidyverse. We will look at the mpg data set.
```{r, echo=FALSE}
library(ggplot2)
mpg
```


Let's start by looking at a scatterplot of engine size (displacement) vs highway mileage (hwy), and answer this question:
Do cars with bigger enginues user more gas? Let's look at the graph of mpg vs engine size
Let's start by looking at a ggplot of the data:

```{r fig.width=7, fig.height=4, echo=FALSE}
library(ggplot2)
ggplot(data = mpg,mapping = aes(x=displ, y=hwy)) +
  geom_point()
```


3.2.4 Exercises

1. Run ggplot(data = mpg). What do you see?
```{r, echo=FALSE}
library(ggplot2)
ggplot(data=mpg)
```

Answer: A blank plot   

2. How many rows are in mpg? How many columns?
```{r, echo=FALSE}
library(ggplot2)
print(paste0("number of rows in mpg = ",nrow(mpg)))
print(paste0("number of columns in mpg = ",ncol(mpg)))
```

What does the drv variable describe? Read the help for ?mpg to find out.

```{r, echo=FALSE}
library(ggplot2)
?mpg
```

The drv variable describes the types of drive systems on the cars:
f = front-wheel drive, r = rear wheel drive, 4 = 4wd

4. Make a scatterplot of hwy vs cyl.
```{r, echo=FALSE}
library(ggplot2)
ggplot(data=mpg) +
  geom_point(mapping = aes(x=hwy, y=cyl))
```

5. What happens if you make a scatterplot of class vs drv? Why is the plot not useful?

```{r, echo=FALSE}
library(ggplot2)
ggplot(mpg, aes(x=class, y=drv)) +
  geom_point() 
```

The plot of class vs drv does not provide useful information.

