---
title: "3 Data Visualization Example"
author: "Russ Conte"
date: "8/4/2018"
output: html_document
---
This chapter will be a look at ggplot2. ggplot2 is part of tidyverse. We will look at the mpg data set.
```{r, echo=FALSE}
library(ggplot2)
mpg
```

#3.2 First Steps
Let's start by looking at a scatterplot of engine size (displacement) vs highway mileage (hwy), and answer this question:
Do cars with bigger enginues user more gas? Let's look at the graph of mpg vs engine size
Let's start by looking at a ggplot of the data:

```{r fig.width=7, fig.height=4, echo=FALSE}
library(ggplot2)
ggplot(data = mpg,mapping = aes(x=displ, y=hwy)) +
  geom_point()
```


###3.2.4 Exercises

1. Run ggplot(data = mpg). What do you see?
```{r}
library(ggplot2)
ggplot(data=mpg)
```

Answer: A blank plot   

2. How many rows are in mpg? How many columns?
```{r}
library(ggplot2)
print(paste0("number of rows in mpg = ",nrow(mpg)))
print(paste0("number of columns in mpg = ",ncol(mpg)))
```

What does the drv variable describe? Read the help for ?mpg to find out.

```{r}
library(ggplot2)
?mpg
```

The drv variable describes the types of drive systems on the cars:
f = front-wheel drive, r = rear wheel drive, 4 = 4wd

4. Make a scatterplot of hwy vs cyl.
```{r}
library(ggplot2)
ggplot(data=mpg) +
  geom_point(mapping = aes(x=hwy, y=cyl))
```

5. What happens if you make a scatterplot of class vs drv? Why is the plot not useful?

```{r}
library(ggplot2)
ggplot(mpg, aes(x=class, y=drv)) +
  geom_point() 
```

The plot of class vs drv does not provide useful information.

#3.3 Aesthetic Mappings

Separate class of the cars by color (note aes in geom_point to get the color):
```{r}
library(ggplot2)
ggplot(mpg, aes(y=hwy, x=displ)) +
         geom_point(aes(color=class))
```

What does the data look like if we map class to size?

```{r}
library(ggplot2)
ggplot(mpg, aes(y=hwy, x=displ)) +
         geom_point(aes(size=class))
```

We can also map class to alpha (transparency), or shape, we follows:

```{r}
library(ggplot2)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))

```

Set the entire data set to the color blue. Note that the color command goes outside of the aes command:

```{r}
library(ggplot2)
ggplot(mpg)+
  geom_point(mapping=aes(y=hwy, x=displ), color="blue")
```

### 3.3.1 Exercises

1. What's wrong with this code? Why aren't the points blue?
```{r}
library(ggplot2)
ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

The points are not blue because the color command goes *outside* of the aes command if it pertains to all points:

```{r}
library(ggplot2)
ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

2. Which variables in mpg are categorical? Which variables are continuous?

```{r}
library(ggplot2)
str(mpg)
```

Categorical: Manufacturer, Model, Trans, Drv, Fl, Class <br>
Continuous: displ, year, cyl, cty, hwy

3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

Three examples of continuous variables:
```{r, warning=FALSE}
library(ggplot2)
ggplot(mpg) +
  geom_point(aes(x=displ, y=cty), color=mpg$hwy)

ggplot(mpg) +
  geom_point(aes(x=displ, y=cty), size=mpg$hwy)

ggplot(mpg) +
  geom_point(aes(x=displ, y=cty), shape=mpg$hwy)
```

An example of discrete variable mapping, this gives multiple errors and halts execution:

4. What happens if you map the same variable to multiple aesthetics?

```{r}
library(ggplot2)
ggplot(mpg) +
  geom_point(aes(x=mpg$displ, y=mpg$cty), color=mpg$displ, shape=mpg$displ, size=mpg$displ)
```

It works, lots of layers on the same plot

5. What does the stroke aesthetic do? What shapes does it work with?

```{r}
library(ggplot2)
ggplot(mpg) +
  geom_point(aes(x=mpg$displ, y=mpg$cty), stroke=mpg$cty)
```

It adds stroke to the shape - the outline surrounding a shape

6 What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)?
This set of commands returns errors and halts execution


