---
title: "3 Data Visualization Example"
author: "Russ Conte"
date: "8/4/2018"
output: html_document
---
This chapter will be a look at ggplot2. ggplot2 is part of tidyverse. We will look at the mpg data set.
```{r, echo=FALSE}
library(ggplot2)
mpg
```

#3.2 First Steps
Let's start by looking at a scatterplot of engine size (displacement) vs highway mileage (hwy), and answer this question:
Do cars with bigger enginues user more gas? Let's look at the graph of mpg vs engine size
Let's start by looking at a ggplot of the data:

```{r fig.width=7, fig.height=4, echo=FALSE}
library(ggplot2)
ggplot(data = mpg,mapping = aes(x=displ, y=hwy)) +
  geom_point()
```


###3.2.4 Exercises

1. Run ggplot(data = mpg). What do you see?
```{r}
library(ggplot2)
ggplot(data=mpg)
```

Answer: A blank plot   

2. How many rows are in mpg? How many columns?
```{r}
library(ggplot2)
print(paste0("number of rows in mpg = ",nrow(mpg)))
print(paste0("number of columns in mpg = ",ncol(mpg)))
```

What does the drv variable describe? Read the help for ?mpg to find out.

```{r}
library(ggplot2)
?mpg
```

The drv variable describes the types of drive systems on the cars:
f = front-wheel drive, r = rear wheel drive, 4 = 4wd

4. Make a scatterplot of hwy vs cyl.
```{r}
library(ggplot2)
ggplot(data=mpg) +
  geom_point(mapping = aes(x=hwy, y=cyl))
```

5. What happens if you make a scatterplot of class vs drv? Why is the plot not useful?

```{r}
library(ggplot2)
ggplot(mpg, aes(x=class, y=drv)) +
  geom_point() 
```

The plot of class vs drv does not provide useful information.

#3.3 Aesthetic Mappings

Separate class of the cars by color (note aes in geom_point to get the color):
```{r}
library(ggplot2)
ggplot(mpg, aes(y=hwy, x=displ)) +
         geom_point(aes(color=class))
```

What does the data look like if we map class to size?

```{r}
library(ggplot2)
ggplot(mpg, aes(y=hwy, x=displ)) +
         geom_point(aes(size=class))
```

We can also map class to alpha (transparency), or shape, we follows:

```{r}
library(ggplot2)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))

```

Set the entire data set to the color blue. Note that the color command goes outside of the aes command:

```{r}
library(ggplot2)
ggplot(mpg)+
  geom_point(mapping=aes(y=hwy, x=displ), color="blue")
```

### 3.3.1 Exercises

1. What's wrong with this code? Why aren't the points blue?
```{r}
library(ggplot2)
ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

The points are not blue because the color command goes *outside* of the aes command if it pertains to all points:

```{r}
library(ggplot2)
ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

2. Which variables in mpg are categorical? Which variables are continuous?

```{r}
library(ggplot2)
str(mpg)
```

Categorical: Manufacturer, Model, Trans, Drv, Fl, Class <br>
Continuous: displ, year, cyl, cty, hwy

3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

Three examples of continuous variables:
```{r, warning=FALSE}
library(ggplot2)
ggplot(mpg) +
  geom_point(aes(x=displ, y=cty), color=mpg$hwy)

ggplot(mpg) +
  geom_point(aes(x=displ, y=cty), size=mpg$hwy)

ggplot(mpg) +
  geom_point(aes(x=displ, y=cty), shape=mpg$hwy)
```

An example of discrete variable mapping, this gives multiple errors and halts execution:

4. What happens if you map the same variable to multiple aesthetics?

```{r}
library(ggplot2)
ggplot(mpg) +
  geom_point(aes(x=mpg$displ, y=mpg$cty), color=mpg$displ, shape=mpg$displ, size=mpg$displ)
```

It works, lots of layers on the same plot

5. What does the stroke aesthetic do? What shapes does it work with?

```{r}
library(ggplot2)
ggplot(mpg) +
  geom_point(aes(x=mpg$displ, y=mpg$cty), stroke=mpg$cty)
```

It adds stroke to the shape - the outline surrounding a shape

6 What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)?
This set of commands returns errors and halts execution

# 3.5 Facets

When doing a facet on one variable (such as class), use facet_wrap:
```{r}
library(ggplot2)
ggplot(mpg) +
  geom_point(aes(x=mpg$displ, y=mpg$cty)) +
  facet_wrap(~class, nrow=2)
```

when using 2 factors, use facet_grid (note this is set to 4 rows, but that can be changed):

```{r}
library(ggplot2)
ggplot(mpg) +
  geom_point(aes(x=mpg$displ, y=mpg$cty)) +
  facet_wrap(~cty, nrow=4)
```

### 3.5.1 Exercises

1. What happens if you facet on a continuous variable, for example milage in the city? (noted as cty in the data)

```{r}
library(ggplot2)
ggplot(mpg) +
  geom_point(aes(x=mpg$displ, y=mpg$cty)) +
  facet_wrap(~cty, nrow=4)
```

2 What do the empty cells in plot with facet_grid(drv ~ cyl) mean? How do they relate to this plot?

```{r}
library(ggplot2)
ggplot(mpg) +
  geom_point(aes(x=mpg$displ, y=mpg$cty)) +
  facet_grid(drv~cyl)
```

No values match the intersection of those variables where the plots are blank

3 What plots does the following code make? What does . do?

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv~.) # facets on the drv variable
```{r}
library(ggplot2)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv~.)
```

This creates facets on the drv variable

```{r}
library(ggplot2)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl) 
```

This plot creates facets on the number of cylinders

Let's remove the . and see what is different:
```{r}
library(ggplot2)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(~cyl)
```

hmmm, no differences that I can see.

# 3.6 Geometric Objects

Left plot in the book:

```{r}
library(ggplot2)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

Right plot in the book:

```{r}
library(ggplot2)
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

How to set up different linetypes by a factor (in this case the drv variable, the car's drivetrain)

```{r}
library(ggplot2)
ggplot(data=mpg) +
  geom_smooth(mapping = aes(x=displ, y=cty, linetype=drv))
```

Also vary by color according to color:

```{r}
library(ggplot2)
ggplot(data=mpg) +
  geom_smooth(mapping = aes(x=displ, y=cty, linetype=drv, color=drv)) +
  geom_point(mapping = aes(x=displ, y=cty, color=drv))
```

Let's look at grouping and how that impacts the plot:

First, no grouping:

```{r}
library(ggplot2)
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy)) 
```

Now group by drive system:

```{r}
library(ggplot2)
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```

```{r}
library(ggplot2)
ggplot(data=mpg)+
  geom_smooth(mapping = aes(x=displ, y=hwy, color=drv),
              show.legend = FALSE)
```

Here is an example of showing multiple geoms on the same plot:

```{r}
library(ggplot2)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))

ggplot(data=mpg,mapping = aes(x=displ, y=cty)) +
  geom_point(mapping = aes(color=class)) +
  geom_smooth()
```

Let's look at how to filter (more on that later) for a specific variable:

```{r}
library(ggplot2)
library(dplyr)
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

## 3.6.1 Exercises

1. What geom would you use to draw a line chart? geom_line()
A boxplot? geom_box()
A histogram? geom_hist()
An area chart? geom_area()

```{r}
library(ggplot2)
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_line(mapping = aes(x=displ, y=hwy))

#boxplot
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_boxplot()

#histogram
ggplot(data = mpg, mapping = aes(x=hwy)) + 
  geom_histogram()

# area cheart
ggplot(data = mpg, mapping = aes(x=hwy, y=displ)) + 
  geom_area()

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```


2. Run this code in your head and predict what it will do, then see what it actually looks like:

```{r}
library(ggplot2)
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```

3. What does show.legend = FALSE do? Removes the legend for the graph
What happens if you remove it? Show the legend, if there is one
Why do you think I used it earlier in the chapter? To show the legend :)

4. What does the se argument to geom_smooth() do?
It shows the standard error for the line/point/etc.
here is se=FALSE:

```{r}
library(ggplot2)
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```

Here is se=TRUE

```{r}
library(ggplot2)
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = TRUE)
```

5. Will these two graphs look different? Why/why not? - they will look the same.

```{r}
library(ggplot2)
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

6. Recreate the R code necessary to generate the following graphs. (A few of these were a very fun challenge!)

```{r}
library(ggplot2)
ggplot(data=mpg) +
  geom_point(mapping=aes(x=displ, y=hwy)) +
  geom_smooth(mapping=aes(x=displ, y=hwy),se=FALSE)
```

```{r}
library(ggplot2)
ggplot(data=mpg) +
  geom_point(mapping=aes(x=displ, y=hwy)) +
  geom_smooth(mapping=aes(x=displ, y=hwy, group=drv),se=FALSE)
```

```{r}
library(ggplot2)
ggplot(data=mpg) +
  geom_point(mapping=aes(x=displ, y=hwy, color=drv)) +
  geom_smooth(mapping=aes(x=displ, y=hwy),se=FALSE)
```

```{r}
library(ggplot2)
ggplot(data=mpg) +
  geom_point(mapping=aes(x=displ, y=hwy, color=drv)) +
  geom_smooth(mapping=aes(x=displ, y=hwy, group=drv, linetype=drv),se=FALSE)
```

Note - I have not yet been able to reproduce the sixth graph in that set.

# 3.7 Statistical Transformations

Interesting question - where do the y-axis values come from in this plot, since they are not in the data set?

```{r}
library(ggplot2)
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

The previous chart can be reproduced using stat_count as well as geom_bar:

```{r}
library(ggplot2)
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

An example where the y value is in the data set alreadey (note the use of stat="identity":

```{r}
library(tidyverse)
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)

ggplot(data=demo) +
  geom_bar(mapping = aes(x=cut, y=freq), stat="identity")

```

It is possible to override the defauls and show proportions (note the use of ..prop..):

```{r}
library(ggplot2)
ggplot(data=diamonds)+
  geom_bar(mapping = aes(x=cut, y=..prop.., group=1))
```

Here is another way to calculate a "stat summary", to summarize y-values for distinct x-values:

```{r}

```

```{r}
library(ggplot2)
ggplot(data=diamonds) +
  stat_summary(
    mapping = aes(x=cut, y=depth),
    fun.ymin = min,
    fun.ymax = max,
    fun.y=median
  )
```

## 3.7.1 Exercises

1. What is the default geom associated with stat_summary()? 
```{r}
library(ggplot2)
?stat_summary
```

stat_summary_bin(mapping = NULL, data = NULL, geom = "pointrange"...
so the default geom is pointrange

How could you rewrite the previous plot to use that geom function instead of the stat function?

```{r}
library(tidyverse)
ggplot(data=diamonds) +
  geom_pointrange(
    mapping = aes(x=cut, y=depth),
    ymin = min(diamonds$depth),
    ymax = max(diamonds$depth)
  )
```

2. What does geom_col() do? How is it different to geom_bar()?

```{r}
library(tidyverse)
?geom_col
```

From the help file: geom_col uses stat_identity: it leaves the data as is.

3. Most geoms and stats come in pairs that are almost always used in concert. Read through the documentation and make a list of all the pairs. What do they have in common?

The names are consistent:
(geom_bar OR geom_col) and stat_count
geom_boxplot and stat_boxplot
geom_bin2d and stat_bin_2d
geom_contour and stat_contour
geom_count and stat_sum
geom_density and stat_density
geom_density_2d and stat_density_2d
geom_hex and stat_bin_hex
(geom_freqpoly OR geom_histogram) and stat_bin
geom_qq and stat_qq
geom_quantile and stat_quantile
geom_ribbon and geom_area
geom_segment and geom_curve
geom_smooth and stat_smooth
geom_label and geom_text
geom_violin and stat_ydensity

4. What variables does stat_smooth() compute? What parameters control its behaviour?

```{r}
library(tidyverse)
?stat_smooth
```

From the help file: Aids the eye in seeing patterns in the presence of overplotting.

5. In our proportion bar chart, we need to set group = 1. Why? In other words what is the problem with these two graphs?

```{r}
library(tidyverse)
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop..))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = ..prop..))
```

The graphs do not show accurate proportions, but using group=1 fixes this problem:

```{r}
library(tidyverse)
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop.., group=1))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = ..prop.., group=1))
```

# 3.8 Position Adjustments

It's possible to add color to a bar chart (or other charts?) using color or fill. Note the color outline on the bars
on the first chart:

```{r}
library(tidyverse)
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
```

Color bars:

```{r}
library(tidyverse)
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```

If color is mapped to a factor with multiple levels, such as Clarity in the Diamonds data set, we get:

```{r}
library(tidyverse)
ggplot(data=diamonds) +
  geom_bar(mapping = aes(x=cut, fill=clarity))
```

Note there are three other options:

1. Position = Identity (note that Identity is the defaul for 2d geoms), it creates overlaps, which is why
alpha is reduced:
```{r}
library(tidyverse)
ggplot(data=diamonds, mapping = aes(x=cut, fill=clarity)) +
  geom_bar(alpha=1/5, position="identity")
```


2. Position = "dodge" instead places overlapping objects <i>beside</i> one another
```{r}
library(tidyverse)
ggplot(data=diamonds, mapping = aes(x=cut, fill=clarity)) +
  geom_bar(alpha=3/5, position="dodge")
```

3. Position = fill - this makes all the bars the same height

```{r}
library(tidyverse)
ggplot(data=diamonds, mapping = aes(x=cut, fill=clarity)) +
  geom_bar(position="fill")
```

4. Position = "jitter" is very good for basic scatter plots. Recall our first scatterplot, it shows 126 points
and we can not tell (for example) if one pair of values has 109 points on the graph, even if the size
of the points are changed:

```{r}
library(tidyverse)
ggplot(mpg,mapping = aes(x=displ, y=hwy))+
  geom_point(mapping = aes(x=displ, y=hwy),size=0.005)
```

but the actual data set has 234 observations. How to show all the observations? Use "jitter" and reduce
the point size:

```{r}
library(tidyverse)
ggplot(mpg,mapping = aes(x=displ, y=hwy))+
  geom_point(mapping = aes(x=displ, y=hwy),position = "jitter", size=0.0005)
```

## 3.8.1 Exercises

1. What is the problem with this plot? How would you improve it?
```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point()
```

One solution to show all the points is to add jitter and reduce the point size:

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point(mapping = aes(x = cty, y = hwy), position="jitter", size=0.0005)
```

2. What parameters to geom_jitter() control the amount of jittering?

```{r}
?geom_jitter
```

The position adjustment, color, size, width of jitter, height of jitter

3. Compare and contrast geom_jitter() with geom_count().

geom_jitter() adds variation to each point, so it's easier to see points
geom_count will count overlapping points

4. What’s the default position adjustment for geom_boxplot()? It is "dodge". Create a visualisation of the mpg dataset that demonstrates it.

(note that ggplot requires values for x an y, this substitutes "" for x, giving the boxplot for city mileage)
```{r}
ggplot(data=mpg) +
  geom_boxplot(mapping = aes(x=mpg$trans, y=cty, group=mpg$trans), position="dodge")
```

Here is an example of the same graph using position = "identity":

```{r}
ggplot(data=mpg) +
  geom_boxplot(mapping = aes(x=mpg$trans, y=cty, group=mpg$trans), position="identity")
```

# 3.9 Coordinate Systems

For example, coord_flip is good for horizontal boxplots. Here is a regular boxplot:

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot()
```

Here is the same boxplot in a horizontal plot:

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot() +
  coord_flip()
```

Another example is coord_quickmap, which plots maps. This is a map of the lower 48 states with <i>incorrect</i> proportions:

```{r}
library(tidyverse)
install.packages("maps")
usa <- map_data("usa")

ggplot(usa, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black")
```

This is a map of the lower 48 states with correct proportions:

```{r}
library(tidyverse)
library("maps")
usa <- map_data("usa")

ggplot(usa, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_quickmap()
```

coord_polar uses polar coordinates, and that can reveal many interesting relationships that
might not be obvious from other charts, such as a bar chart:

```{r}
bar <- ggplot(data = diamonds) + 
  geom_bar(
    mapping = aes(x = cut, fill = cut), 
    show.legend = FALSE,
    width = 1
  ) + 
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)

bar + coord_flip()
bar + coord_polar()
```

## 3.9.1 Exercises

1. Turn a stacked bar chart into a pie chart using polar_coord()

```{r}
library(tidyverse)
bar1 <- ggplot(data=diamonds, mapping = aes(x=cut, fill=clarity)) +
  geom_bar(alpha=4/5, position="identity")

bar1+coord_polar()
```

2. What does labs() do? Read the documentation.

```{r}
?labs
```

Good labels will:
1. Ensure x and y axes are easily understood
2. Plot title will let people know the subject/topic of the plot
3. Captions can be used to help the reader understand the plot in more detail

3. What’s the difference between coord_quickmap() and coord_map()?

Here is coord_quickmap():

```{r}
library(tidyverse)
library("maps")
usa <- map_data("usa")

ggplot(usa, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_quickmap()
```

Here is an example of quickmap that shows the lower 48 states:

```{r}
library(tidyverse)
library(evaluate)
states <- map_data("state")
usamap <- ggplot(states, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black")
usamap + coord_quickmap()

```

4. What does the plot below tell you about the relationship between city and highway mpg? Why is coord_fixed() important? What does geom_abline() do?

```{r}
install.packages("ggplot2")
library(ggplot2)
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() + 
  geom_abline() +
  coord_fixed()
```

This shows that highway and city mileage tend to go up together. As one goes up, the other goes up, too.
geom_abline() draws a straight line that shows the straight line relationship between highway and city mileage.

Why is coord_fixed() important?